Covariate analysis of ROSMAP RNASeq with Quantile and SVA normalisation

### 2016-09-07 code/Rmd/ROSMAP_RNASeq_Qntl_SVA.Rmd
Added markdown to process FPKM from ROSMAP using quantile normalisation and SVA

---
title: "Covariate analysis of ROSMAP RNASeq"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Qntl (Based on cogdx)')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570291';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of ROSMAP RNASeq data';

thisFileName <- 'ROSMAP_RNASeq_Qntl.Rmd'

@@

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=15, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (custom design)
Fitting a linear model with cogdx, Batch, RINcontinuous, and pmi 
```{r iterative.normalisation, cache=TRUE, results='asis', echo=FALSE}
postAdjustCovars <- c('cogdx', 'Batch', 'RINcontinuous', 'pmi')

writeLines(paste('Using following covariates in the final model',paste(postAdjustCovars,collapse=','),'as fixed effects',sep=' '))

@@

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
rownames(DESIGN) <- rownames(COVARIATES)
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
                                              DESIGN,
                                              'all adjusted design',
                                              isKeyPlot=TRUE)
```

```{r residualSigCovars.manual.ALL, fig.width=15, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding cogdx back to the residuals
```{r addVarsBack}
varsToAddBackIn <- c("cogdx1",  "cogdx2", "cogdx3", "cogdx4", "cogdx5", "cogdx6")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + t(data.matrix(DESIGN[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'EnsemblGeneID') %>%
  tidyr::separate(col = EnsemblGeneID, into = c('ensembl_gene_id','position'), sep = '\\.')
```

Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION[, !(colnames(RESIDUAL.GENE_EXPRESSION) %in% c('ensembl_gene_id','position'))]))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual gene expression
```{r decompse.normalise.data3, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1:2)], scale.=T)

@@

COVARIATES.tmp = data.matrix(COVARIATES)
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
```

@@

@@

kable(tmp)
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP RNASeq Qntl (Based on cogdx)", parentId = parentId)
CODE <- synStore(CODE)

@@

# Store covariates
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'FPKM Normalised Adjusted Residual Expression (cogdx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

>>>>

---
title: "Covariate analysis of ROSMAP RNASeq with Quantile and SVA normalisation"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(knitr)

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Qntl_SVA.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Qntl SVA (Based on cogdx)')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(Rcolorbrewer)

library(sva)

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570291';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of ROSMAP RNASeq data with quantile normalisation and SVA adjustments';

thisFileName <- 'ROSMAP_RNASeq_Qntl_SVA.Rmd'

@@

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=15, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising FPKM with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'cogdx'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

@@

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  rownames(DESIGN) <- rownames(COVARIATES)
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
                                                DESIGN,
                                                'all adjusted design',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('cogdx',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL1 = DM1$design

## Calculate the number of surrogate variables
# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
# SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
# NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
# SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
# SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]

# # Permute and calculate the permuted singular values
# PERM.SING.VALS <- foreach (i = 1:100,
#          .combine = rbind,
#          .packages = c('limma','CovariateAnalysis'),
#          .export = c('PROCESSED_COUNTS','MODEL1')) %dopar% {
#   PERM.COUNTS <- t(apply(PROCESSED_COUNTS$filteredExprMatrix, 1, sample, replace = FALSE))
#   # Estimate voom weights
#   VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
#   
#   # Fit linear model using new weights and new design
#   VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
#   
#   # Residuals after normalisation
#   RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
#   
#   PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
#   PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
# }
# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])

num.sv2 = num.sv(EXPR, MODEL1, method = 'be', seed = 123456)
NUM.SV = num.sv2 #min(num.sv1, num.sv2)

## Compute surrogate variables 
SURR.VAR = sva(EXPR, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

# Fit linear model using estimated surrogate variables and postadjust covariates
FIT = lmFit(EXPR, design = cbind(MODEL1, SURR.VAR))

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,EXPR)
```

### Residual calculation
Calculate weighted residuals and add back "cogdx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  FIT$coefficients[,varsToAddIn] %*% t(FIT$design[,varsToAddIn])
```

Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual gene expression
```{r decompse.normalise.data3, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Batch), shape=as.factor(cogdx), size=as.numeric(age_death)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= Sampleid), size=4, hjust=1)
p
```

@@

RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id.position') %>%
  tidyr::separate(ensembl_gene_id.position, c('ensembl_gene_id', 'position'), sep = '\\.')

@@

kable(tmp)
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP RNASeq Qntl SVA (Based on cogdx)", parentId = parentId)
CODE <- synStore(CODE)

@@

# Store design matrix
MODEL1 = rownameToFirstColumn(MODEL1, 'Sampleid_batch')
write.table(MODEL1, file = 'designMatrix.tsv', sep = '\t', row.names=F, quote=F)
DES_OBJ = File('designMatrix.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DES_OBJ = synStore(DES_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store surrogate variables 
rownames(SURR.VAR) = rownames(MODEL1)
SURR.VAR = rownameToFirstColumn(SURR.VAR, 'Sampleid_batch')
write.table(SURR.VAR, file = 'SurrogateVariable.tsv', sep = '\t', row.names=F, quote=F)
SV_OBJ = File('SurrogateVariable.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
SV_OBJ = synStore(SV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residualised expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'FPKM Normalised Adjusted Residual Expression (cogdx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
